---
title: "Compartmental Diagram Interactive Session"
execute:
    echo: true
    warning: false
editor:
  markdown:
    wrap: sentence
metadata-files:
    - metadata/matthewferrari.yml
---

In this session, you will start with a simple SIR model diagram and modify it to represent an intervention or transmission process that is interesting to your group.
The goal is not to make the most complicated model possible.
Instead, the goal is to make each modeling choice explicit: what compartments are needed, what flows connect them, and what assumptions those flows imply.

## Learning goals

By the end of this session, you should be able to:

1. understand how to translate a public-health intervention into a compartment model structure;
2. identify the states, flows, and parameters in that structure;
3. explain which assumptions are introduced by adding or removing arrows; and
4. describe the expected qualitative effect of the intervention on epidemic dynamics.

## Setup

We will use the `diagram` package to draw box-and-arrow compartment diagrams, as in [R Session 02](./r-session-02.qmd), and the `purrr` package to create uniform compartment boxes.

```{r}
library(diagram)
library(purrr)
```

## Starting point: a simple SIR diagram

The basic SIR model has three compartments:

- $S$: susceptible individuals;
- $I$: infectious individuals; and
- $R$: recovered or removed individuals.

```{r}
#| echo: true
#| column: body
#| out-width: 80%
#| code-fold: true

elpos <- rbind(
  S = c(1, 1),
  I = c(2, 1),
  R = c(3, 1)
)

elpos[, 1] <- (2 * elpos[, 1] - 1) / 6
elpos[, 2] <- 0.5

fromto <- rbind(
  SI = c(1, 2),
  IR = c(2, 3)
)

op <- par(mar = c(1, 1, 1, 1))
diagram::openplotmat(asp = 0.35)

for (i in seq_len(nrow(fromto))) {
  diagram::straightarrow(
    to = elpos[fromto[i, 2], ],
    from = elpos[fromto[i, 1], ],
    lwd = 2,
    arr.pos = 0.65,
    arr.length = 0.5
  )
}

purrr::walk(
  c("S", "I", "R"),
  .f = function(.x) {
    diagram::textrect(
      elpos[.x, ],
      0.08,
      0.10,
      lab = .x,
      box.col = gray(0.7),
      shadow.col = gray(0.4),
      shadow.size = 0.01,
      cex = 2
    )
  }
)

text(mean(elpos[c("S", "I"), 1]), 0.62, expression(lambda), cex = 1.8)
text(mean(elpos[c("I", "R"), 1]), 0.62, expression(gamma), cex = 1.8)

par(op)
```

For the standard SIR model, infection moves people from $S$ to $I$ at rate $\lambda$, and recovery or removal moves people from $I$ to $R$ at rate $\gamma$.

<div class="exercise">

### Exercise 1: Choose one intervention or modeling feature that your group wants to represent. Then change the SIR diagram to include it.

</div>
